---
title: "ggsubcat: Sub-categorical color scales and guides with ggplot2"
output: rmarkdown::html_vignette
author: Yichen Wang
vignette: >
  %\VignetteIndexEntry{ggsubcat: Sub-categorical color scales and guides with ggplot2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  fig.align = "center"
)
```

Admitting that human eyes cannot easily distinguish too many colors when we have many categories to be colored on a plot, a clever way is always in need. For the specific scenario where only a few categories exist while they can be sub-devided into many more sub-categories, this package proposes to distinguish the sub-categories within one main category by luminance (lightness) and chroma (intensity of color), while distinguishing main categories by different colors you can tell by words.

This package is dedicated to keep the grammer of graphics philosophy of ggplot2, so you can make the sub-categorical coloring happen as simply as adding an additional aesthetic and a scale to the call chain of ggplot2, with all other flexibility retained.

```{r setup}
library(ggsubcat)
library(ggplot2)
library(cowplot)
```

## Example data simulation

First, let's begin with some simulated data. Here we bring totally 900 data points grouped into 3 main categories, each divided into 3 sub-categories. The x and y positions are selected to reflect the relationship between the information, which you will see in the scatter plot later.

```{r data}
set.seed(27)
df <- data.frame(
    x = c(rnorm(100, 1, 0.3), rnorm(100, 2, 0.3), rnorm(100, 3, 0.3),
          rnorm(100, 7, 0.3), rnorm(100, 8, 0.3), rnorm(100, 9, 0.3),
          rnorm(100, 4, 0.3), rnorm(100, 5, 0.3), rnorm(100, 6, 0.3)),
    y = c(rnorm(100, 1, 0.3), rnorm(100, 2, 0.3), rnorm(100, 1, 0.3),
          rnorm(100, 1, 0.3), rnorm(100, 2, 0.3), rnorm(100, 1, 0.3),
          rnorm(100, 5, 0.3), rnorm(100, 6, 0.3), rnorm(100, 5, 0.3)),
    main = factor(c(
        rep("A", 300),
        rep("B", 300),
        rep("C", 300)
    )),
    sub = factor(c(
        rep("a1", 100), rep("a2", 100), rep("a3", 100),
        rep("b1", 100), rep("b2", 100), rep("b3", 100),
        rep("c1", 100), rep("c2", 100), rep("c3", 100)
    ))
)
```

This should be actually equivalent to the package-delivered dataset `subcatPoints`

```{r data2}
data("subcatPoints")
```

## Coloring points with sub-categorical information

Scatter plot case. We show that we color the dots by the main categories or by the sub-categories first.

```{r plot, fig.height=3, fig.width=7}
p.main <- ggplot(df, aes(x, y, color = main)) +
    geom_point()
p.sub <- ggplot(df, aes(x, y, color = sub)) +
    geom_point()
plot_grid(p.main, p.sub, ncol = 2)
```

As you can see on the left panel, the dots are forming three main populations, while on the right panel, each main population is divided into three sub-populations, closely placed. This is very similar to the scenario in bioinformatics analysis, where course annotation of cell types can be finely annotated to sub-types, and researchers need to somehow color the dots on the low-dimensional embedding plot to visualize the cell populations.

Then with `scale_colour_subcat()`, we can color the dots by sub-categories, with the main category information retained.

```{r plot2, fig.width=5, fig.height=4}
ggplot(df, aes(x, y, color = sub, color.group = main)) +
    geom_point() +
    scale_colour_subcat()
```

## `fill` aesthetic can also be used

Similarly, with a bar plot where the aesthetic `fill` is to be used, we color the barsby the main categories or the sub-categories first.

```{r plot3, fig.width=7, fig.height=3}
b.main <- ggplot(df, aes(x = sub, y = y, fill = main)) +
    geom_bar(stat = "identity")
b.sub <- ggplot(df, aes(x = sub, y = y, fill = sub)) +
    geom_bar(stat = "identity")
plot_grid(b.main, b.sub, ncol = 2)
```

Then with `scale_fill_subcat()`, we can color the bars by sub-categories, with the main category information retained.

```{r plot4, fig.width=5, fig.height=4}
ggplot(df, aes(x = sub, y = y, fill = sub, fill.group = main)) +
    geom_bar(stat = "identity") +
    scale_fill_subcat()
```
